### 2020-10-27 doc/manuscript/Dow_manuscript.Rmd
big-picture structuring of intro

#Table numbers

#Figure numbers
schematic=1 #Based on D'Orangeville Figure 1
growth_pheno_by_xylem = 2  #if we include more sites, we'll make this multi-panel
rel_growth = "2a" #Based on D'Orangeville Figure 3.a), but with relative growth
cum_growth = "2b" #Similar to D'Orangeville Figure 4.a), but with cumulative relative growth
climwin = 3
pheno_Tsensitivity = 4
model_summary = 5
DOYtiming = 6
#SI:
#tables

@@

As global atmospheric greenhouse gas levels are rising, and little meaningful policy is being implemented in response, our planet is expected to surpass the 2.5 degree C warming mark that the Paris accord pledged to avoid. The absence of human intervention increases our reliance on natural systems to soften the blow of climate change. Plant life, especially forests, make up a large portion of these natural systems, offering many valuable ecosystem services. For one, they house an astonishing 80% of terrestrial species and are capable of removing the greenhouse gas CO2 from the atmosphere and storing it in the trees or soil. Forests sequester about one-third of all anthropogenic CO2 and store ~45% of all terrestrial carbon (Bonan, 2008). However, their future under the current era of rapid change is uncertain (Friedlingstein et al, 2006).

Reducing this uncertainty requires better understanding of the drivers of tree growth (Bugmann et al, 2019). Tree growth is sensitive to climatic factors like precipitation, temperature, and cloudiness in early growing season (Helcoski et al, 2019; Ciais, 2003; Goulden, 1996), as well as other physical factors including nitrogen content of soil, increasing atmospheric CO2, deposition of atmospheric pollutants in acid rain (Mathias & Thomas, 2018), and invasion by pests or pathogens. Understanding how growth is influenced by these factors is essential for developing accurate models to inform forest managers and policy makers about the role of future forests in climate change mitigation efforts.

In recent decades, an increase in average Spring temperature has been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006). The effects of this Spring warming are most apparent when examining leaf phenology shifts, which have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018), leading to many critical observations. As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by later leaf senescence, meaning a potential shift in growing season timing or length (Ibáñez et al, 2010; Jeong et al., 2011). The timing of this Spring warming also has a significant effect on leaf phenology response (Crabbe et al, 2016).

Shifts in leaf phenology timing could be accompanied by changes in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;). However, the fine-scale responses of temperate deciduous tree growth to a warming Spring are not commonly recorded due to the increased effort required (). Many observations of climate induced phenological shifts are based on GIS (), eddy-covarience measurements (), or other broad scale measurements of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review). Further to this point, a recent publication by D'Orangeville et al () found that leaf and wood phenology  traits, along with their apparent relationships, are dissimilar among trees of different wood types (ring-porous and diffuse-porous). For example, they observed that bud burst was occurring around the same time as 25% total annual growth, and 75% leaf development coincided with 50% total annual growth in ring-porous trees, while leaf phenology and stem growth phenology were decoupled in diffuse-porous trees.

The significance of wood type in regards to wood and leaf phenology likely stems from evolutionary differences and adaptations to local climates where these trees can be found. Ring-porous trees have large xylem cells which are prone to embolism during the winter. These cells are repaired before leaf development, explaining why the 25% yearly growth milestone is reached around the same time as bud burst. Diffuse-porous xylem cells are smaller, and less prone to embolism during the Winter, meaning they do not need to be repaired in order to restore hydraulic functionality before beginning the leaf development process. Results from the paper by D’Orangeville et al. (in review) have shown that ring-porous trees reach 25 and 50% of their total yearly growth during Spring, while diffuse-porous trees reach these milestones in the Summer, but both reached 75% total annual growth around the same time. --Consequently, the Summer focused diffuse-porous trees are exposed to higher temperatures and drought risk than ring-porous trees.-- If leaf development is sped up when Spring temperatures are warmer, we should seek to understand whether this relationship holds, and how it potentially varies among trees of different wood-types.

@@

(May remove... not sure where to work in...or if it needs to be in) Another study found that the velocity of leaf development is correlated to bud burst timing, such that the later in the Spring bud burst occurred, the faster leaves developed and vice versa (Klosterman & Hufkins & Richardson, 2018).

While the afformentioned broad scale studies have shown a potential relationship between total tree growth and temperature, timing of intraannual stem-growth is extremely important as well. The timing of tree growth has profound impacts on ecosystem interactions, as well as dictates periods of highest drought sensitivity in temperate forests (Foster, 2014; ). Species have co-evolved to utilize the phenological stages of trees at specific times in the year. If the timing of intraannual phenology traits were to shift, that could harm the species depending on the trees (). Additionally, if a drought were to occur when trees were at their peak growth rates, the resulting decrease in that year's NPP would reduce the total amount of carbon sequestered, contributing to a climate feedback loop further exacerbating these potential shifts (source?).

While not entirely impossible, it is difficult to record intraannual growth traits or compare interspecific responses to Spring warming using leaf-phenology or other indirect observations. However, some traditional methods of direct observation are labor intensive, and very time consuming (Source about cell sampling). A solution to this is the use of dendrometer bands. Dendrometer bands allow a look at individual trees’ intraannual growth at a fine temporal scale (source?). Using many measurements taken within a year, we can model the growth of individual trees and extract data on growth milestones and traits for interspecific, intraspecific, and temporal comparison (McMahon & Parker, 2014). This allows the determination of variable responses of key tree species within a forest, meaning a better understanding of future growth under projected climate conditions. Here, we use 15 (5 @ Harvard Forest, 10 @ SCBI) years of biweekly dendrometer band measurements for 19 (15 @ Harvard Forest, 4 @ SCBI) species to characterize intraannual stem-growth phenology within two forest dynamics plots in the Eastern USA, calculating for each tree in each year where data was available: the day of year (DOY) where 25, 50, and 75% annual growth was achieved; the maximum growth rate; the DOY where maximum growth rate was achieved; the total growth; and season length (75%-25% DOY). Correlation between these variables and Spring temperatures over the 15 year study period (1998-2003; 2011-2019) were assessed using a linear mixed model in a heriarchal bayesian framework.

We hypothesized that (1) stem-growth milestones would be reached earlier in warmer years as compared to cooler years and (2) both wood types would increase total growth and max growth rate in response to warmer Spring temperatures.

@@

1. Model was fit to a random sample of 20 out of 106 tags to speed up computation. *(This isn't the case anymore, I fit it to all tag years)*
1. Climwin mean temperatures are recentered at 65F to help sanity check intercepts and slopes. 65F was chosen because it is a "typical" climwinmean value *(Changed to 16 C based on updated climwin windows)*
1. The multivariate model for $(DOY_{25}, DOY_{50}, DOY_{75})$ has:
    1. Fixed effects: The interaction between wood type (diffuse vs ring porous) and climwin mean
    1. Random effects: tag number (no species just yet)

@@

Content to incorporate: N. American strategies have conservative strategies when it comes to phenology, as historically they’ve been subject to more spring frosts. Thus, climate change is having less impact (Zohner et al., 2020); Leaf velocity discussion? - - RP trees BB = 25% and 75% leaf development = 50% growth (D'Orangeville). Are DP trees leafing up earlier? If yes, how much earlier? Similar to 25,50,75% DOY retreat?

>>>>

#Table numbers
hypothesis_table=1

#Figure numbers
schematic=1 #illustrates parameters
growth_pheno_by_xylem = 2  #if we include more sites, we'll make this multi-panel
rel_growth = "2a" #Based on D'Orangeville Figure 3.a), but with relative growth
cum_growth = "2b" #Similar to D'Orangeville Figure 4.a), but with cumulative relative growth
climwin = 3
pheno_Tsensitivity = 4
model_summary = 5
DOYtiming = 6
#SI:
#tables

@@

#### (1. forests are critical for climate change regulation, so we need to understand better how they're responding to climate change)
As global atmospheric greenhouse gas levels are rising, and little meaningful policy is being implemented in response, our planet is expected to surpass the 2.5 degree C warming mark that the Paris accord pledged to avoid. The absence of human intervention increases our reliance on natural systems to soften the blow of climate change. Plant life, especially forests, make up a large portion of these natural systems, offering many valuable ecosystem services. For one, they house an astonishing 80% of terrestrial species and are capable of removing the greenhouse gas CO2 from the atmosphere and storing it in the trees or soil. Forests sequester about one-third of all anthropogenic CO~2~ and store ~45% of all terrestrial carbon (Bonan, 2008). However, their future under the current era of rapid change is uncertain (Friedlingstein et al, 2006).

Reducing this uncertainty requires better understanding of the drivers of tree growth (Bugmann et al, 2019). Tree growth is sensitive to climatic factors like precipitation, temperature, and cloudiness in early growing season (Helcoski et al, 2019; Ciais, 2003; Goulden, 1996), as well as other physical factors including nitrogen content of soil, increasing atmospheric CO~2~, deposition of atmospheric pollutants in acid rain (Mathias & Thomas, 2018), and invasion by pests or pathogens. Understanding how growth is influenced by these factors is essential for developing accurate models to inform forest managers and policy makers about the role of future forests in climate change mitigation efforts.

#### (2. spring temperatures are rising, and this is affecting leaf phenology and carbon cycling)
In recent decades, an increase in average Spring temperature has been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006).** The effects of this Spring warming are most apparent when examining leaf phenology shifts, *which have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018), leading to many critical observations*. As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by later leaf senescence, meaning a potential shift in growing season timing or length (Ibáñez et al, 2010; Jeong et al., 2011). The timing of this Spring warming also has a significant effect on leaf phenology response (Crabbe et al, 2016).

#### (3. However, we don't know how warming springs affect stem-growth, which matters because the C allocated to stem growth will be retained in the ecosystem for decades to millenia )
Shifts in leaf phenology timing could be accompanied by changes in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;). However, the fine-scale responses of temperate deciduous tree growth to a warming Spring are not commonly recorded due to the increased effort required (). Many observations of climate induced phenological shifts are based on GIS (), eddy-covarience measurements (), or other broad scale measurements of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review).

Tree-ring records give some hints that warmer springs may increase annual growth (Helcoski et al. 2019), but tree-rings tell us nothing about the timing of that growth.  Timing of intraannual stem-growth is extremely important as well.

@@

#### (4. We also know little about differences across species or functional groups, which matters because it will affect how different species fare under climate change )
A recent publication by D'Orangeville et al () found that leaf and wood phenology  traits, along with their apparent relationships, are dissimilar among trees of different wood types (ring-porous and diffuse-porous). For example, they observed that bud burst was occurring around the same time as 25% total annual growth, and 75% leaf development coincided with 50% total annual growth in ring-porous trees, while leaf phenology and stem growth phenology were decoupled in diffuse-porous trees.

The significance of wood type in regards to wood and leaf phenology likely stems from evolutionary differences and adaptations to local climates where these trees can be found. Ring-porous trees have large xylem cells which are prone to embolism during the winter. These cells are repaired before leaf development, explaining why the 25% yearly growth milestone is reached around the same time as bud burst. Diffuse-porous xylem cells are smaller, and less prone to embolism during the Winter, meaning they do not need to be repaired in order to restore hydraulic functionality before beginning the leaf development process. Results from the paper by D’Orangeville et al. (in review) have shown that ring-porous trees reach 25 and 50% of their total yearly growth during Spring, while diffuse-porous trees reach these milestones in the Summer, but both reached 75% total annual growth around the same time. If leaf development is sped up when Spring temperatures are warmer, we should seek to understand whether this relationship holds, and how it potentially varies among trees of different wood-types.

#### (5. Here, we...)
Using many measurements taken within a year, we can model the growth of individual trees and extract data on growth milestones and traits for interspecific, intraspecific, and temporal comparison (McMahon & Parker, 2014). This allows the determination of variable responses of key tree species within a forest, meaning a better understanding of future growth under projected climate conditions. Here, we use biweekly dendrometer band measurements (5 years @ Harvard Forest, 10 @ SCBI) for dominant tree species at each site (15 @ Harvard Forest, 4 @ SCBI)  to characterize intraannual stem-growth phenology within two forest dynamics plots in the Eastern USA, calculating for each tree in each year where data was available: the day of year (DOY) where 25, 50, and 75% annual growth was achieved; the maximum growth rate; the DOY where maximum growth rate was achieved; the total growth; and season length (75%-25% DOY) (Fig. `r schematic`). Correlation between these variables and Spring temperatures over the 15 year study period (1998-2003; 2011-2019) were assessed using a linear mixed model in a heriarchal bayesian framework.

We test the hypotheses that (1) stem-growth milestones are reached earlier in years with warmer early springs, but (2) maximum growth rates are independent of spring temperatures, and as a result (3) total annual growth increases in response to warmer spring temperatures (Table `r hypothesis_table`).

```{r hypothesis_table, eval=TRUE, echo=FALSE, warning=FALSE}
library(knitr)  
hypothesis_table <- read.csv(file = "tables_figures/hypothesis_table.csv")
knitr::kable(hypothesis_table, caption = "Table 1. Hypothesis table. (NEEDS FORMATTING...obviously!)")
```

@@

1. The multivariate model for $(DOY_{25}, DOY_{50}, DOY_{75})$ has:
    1. Fixed effects: The interaction between wood type (diffuse vs ring porous) and climwin mean
    1. Random effects: tag number (no species just yet)

@@

### (Content to incorporate)

N. American strategies have conservative strategies when it comes to phenology, as historically they’ve been subject to more spring frosts. Thus, climate change is having less impact (Zohner et al., 2020); Leaf velocity discussion? - - RP trees BB = 25% and 75% leaf development = 50% growth (D'Orangeville). Are DP trees leafing up earlier? If yes, how much earlier? Similar to 25,50,75% DOY retreat?

Another study found that the velocity of leaf development is correlated to bud burst timing, such that the later in the Spring bud burst occurred, the faster leaves developed and vice versa (Klosterman & Hufkins & Richardson, 2018).

The timing of tree growth has profound impacts on ecosystem interactions, as well as dictates periods of highest drought sensitivity in temperate forests (Foster, 2014; D’Orangeville et al. in review). Species have co-evolved to utilize the phenological stages of trees at specific times in the year. If the timing of intraannual phenology traits were to shift, that could harm the species depending on the trees (). Additionally, if a drought were to occur when trees were at their peak growth rates, the resulting decrease in that year's NPP would reduce the total amount of carbon sequestered, contributing to a climate feedback loop further exacerbating these potential shifts (source?).
